---
title: "CE2: Working with Master Plan Planning Subzone Data"
subtitle: "In-class Exercise 2"
author: "Kendrick Teo"
date: "2024-08-26"
date-modified: "last-modified"

execute: 
  eval: true
  echo: true
  freeze: true
---

## CE2.1 Working with Masterplan 2014 Subzone Data

1. Create a subfolder called *data* in In-class_Ex02 folder

2. Download and load both the ESRI shapefile and `kml` file from *Master Plan 2014 Subzone Boundary (Web)* from the portal.

3. Write a code chunk to import *Master Plan 2014 Subzone Boundary (Web)* as SF simple features DataFrame.

```{r}
pacman::p_load(sf, tmap, tidyverse, ggstatsplot)

mp_subzone_14_shp <- st_read(dsn="data/MasterPlan2014SubzoneBoundaryWebSHP", layer="MP14_SUBZONE_WEB_PL")
class(mp_subzone_14_shp)

#! eval:false
# Doesn't work!
tryCatch({mp_subzone_kml <- st_read("data/MasterPlan2014SubzoneBoundaryWebKML.kml")}, error=function(e) {
  print(e)
  print("Cannot load!")
  })

# So we need to write our own KML file.
st_write(mp_subzone_14_shp, "data/WORKINGMasterPlan2014SubzoneBoundaryWebKML.kml", delete_dsn=TRUE)
# delete_dsn replaces the file with our new one if its already exists
```

## CE2.2 Working with Masterplan 2019 Subzone Data

```{r}
mp_subzone_19_shp <- st_read(dsn="data/MPSZ-2019", layer="MPSZ-2019")
mp_subzone_19_kml <- st_read("data/MasterPlan2019SubzoneBoundaryNoSeaKML.kml")
```

## CE2.3 Transforming coordinate data from WGS84 to SVY21

```{r}
mp_subzone_19_shp <- mp_subzone_19_shp %>% st_transform(crs = 3414)
st_crs(mp_subzone_19_shp)
```


## CE2.4 Working with Population Data

```{r}
population_2023 <- read_csv("data/respopagesextod2023/respopagesextod2023.csv")
print("Columns of population_2023")
colnames(population_2023)
print("Head of population_2023")
head(population_2023)
```

## CE2.5 Group data by planning area and subzone

```{r}
population_2023_pivot <- population_2023 %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP`=sum(`Pop`)) %>%
  ungroup() %>%
  pivot_wider(names_from=AG, values_from=POP)
colnames(population_2023)
# Read the data structure carefully before proceeding!
```

## CE2.6 