---
title: "H1: Geospatial Data Wrangling with R"
subtitle: "Hands-On Exercise 1"
date: "2024-08-17"         # Published date
last-modified: "2024-08-17" # Modified date
---

# H1.1 Overview

As adapted from the [course website](https://r4gdsa.netlify.app/chap01learning-outcome):

Geospatial Data Science is a process of importing, wrangling, integrating, and processing geographically referenced data sets. In this hands-on exercise, you will learn how to perform geospatial data science tasks in R by using **sf** package.

By the end of this hands-on exercise, you should acquire the following competencies:

-   installing and loading [sf](https://r-spatial.github.io/sf/) and [tidyverse](https://www.tidyverse.org/) packages into R environment,

-   importing geospatial data by using appropriate functions of **sf** package,

-   importing aspatial data by using appropriate function of **readr** package,

-   exploring the content of simple feature data frame by using appropriate **Base R** and **sf** functions,

-   assigning or transforming coordinate systems by using using appropriate *sf* functions,

-   converting an aspatial data into a sf data frame by using appropriate function of **sf** package,

-   performing geoprocessing tasks by using appropriate functions of **sf** package,

-   performing data wrangling tasks by using appropriate functions of **dplyr** package and

-   performing Exploratory Data Analysis (EDA) by using appropriate functions from **ggplot2** package.

> Note: Students are encouraged to read the reference guide of each function, especially the input data requirements, syntaxt and argument option before using them.

# H1.2 Data acquisition

To complete this exercise, the following datasets were downloaded and extracted into `Hands-on_Ex01/data/`:

-   Master Plan 2014 Subzone Boundary (Web) from [data.gov.sg](https://data.gov.sg/)
-   Pre-Schools Location from [data.gov.sg](https://data.gov.sg/)
-   Cycling Path from [LTADataMall](https://www.mytransport.sg/content/mytransport/home/dataMall.html)
-   Latest version of Singapore Airbnb listing data from [Inside Airbnb](http://insideairbnb.com/get-the-data.html)

# H1.3 Getting Started

In this hands-on exercise, two R packages were used. They are:

-   **sf** for importing, managing, and processing geospatial data, and

-   **tidyverse** for performing data science tasks such as importing, wrangling and visualising data.

Tidyverse consists of a family of R packages. In this hands-on exercise, the following packages have been used:

-   **readr** for importing csv data,

-   **readxl** for importing Excel worksheet,

-   **tidyr** for manipulating data,

-   **dplyr** for transforming data, and

-   **ggplot2** for visualising data

    ```{r}
    pacman::p_load(sf, tidyverse)
    ```

# H1.4 Importing Geospatial Data

Geospatial data was imported into R using [*st_read()*](https://r-spatial.github.io/sf/reference/st_read.html) of **sf** package:

-   `MP14_SUBZONE_WEB_PL`, a polygon feature layer in ESRI shapefile format,

-   `CyclingPath`, a line feature layer in ESRI shapefile format, and

-   `PreSchool`, a point feature layer in kml file format.

## H1.4.1 Importing polygon feature data in shapefile format

The code chunk below uses *st_read()* function of **sf** package to import the `MP14_SUBZONE_WEB_PL` shapefile into R as a polygon feature data frame.

```{r}
mp_subzone = st_read(dsn="data/geospatial/MasterPlan2014SubzoneBoundaryWebSHP", layer="MP14_SUBZONE_WEB_PL")
```

There are a total of 323 multipolygon features and 15 fields in `mpsz` simple feature data frame. `mpsz` is in **svy21** projected coordinates systems. The bounding box provides the x extend and y extend of the data.

## H1.4.2 Importing polyline feature data in shapefile form

Using the same method as above, we now import the `CyclingPath` shapefile into R as a line feature data frame.

```{r}
cycling_paths = st_read(dsn="data/geospatial/CyclingPath_Jul2024", layer="CyclingPathGazette")
```

## H1.4.3 Importing GIS data in kml format

```{r}
preschool_data = st_read("data/geospatial/PreSchoolsLocation.kml")
```

# H1.5 Checking the Content of a Simple Feature DataFrame

## H1.5.1 Working with *st_geometry()*

```{r}
st_geometry(mp_subzone)
```

## H1.5.2 Working with *glimpse()*

```{r}
glimpse(mp_subzone)
```

## H1.5.3 Working with *head()*

```{r}
head(mp_subzone, n=10)
```

# H1.6 Plotting the Geospatial Data

```{r}
max.plot = 15
plot(mp_subzone)
```