---
title: "H2: Thematic Mapping and GeoVisualisation with R"
subtitle: "Hands-On Exercise 2"
author: "Kendrick Teo"
date: "2024-08-19"
date-modified: "last-modified"

execute: 
  eval: true
  echo: true
  freeze: true
---

# H2.1 Overview

While thematic mapping uses map symbols to visualise geographic properties such as population and temperature, geovisualisation is a subset of thematic mapping, where a pseudocolour is used to correspond with these geographic properties.

# H2.2 Getting Started

`tmap` contains the packages needed to perform geovisualisation, so we need to import that along with the usual packages we use.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

# H2.3 Importing Data into R

The datasets to be used are:

1. Master Plan 2014 Subzone Boundary (Web)(`MP14_SUBZONE_WEB_PL`) in ESRI shapefile format
```{r}
mp_subzone = st_read(dsn="../data/geospatial/MasterPlan2014SubzoneBoundaryWebSHP", layer="MP14_SUBZONE_WEB_PL")
mp_subzone
```

2. Singapore Residents by Planning Area / Subzone, Age Group, Sex and Type of Dwelling, June 2011-2020 (aspatial)

```{r}
population_data = read_csv("../data/aspatial/respopagesextod2011to2020/respopagesextod2011to2020.csv")
head(population_data)
```

# H2.3.4 Data Preparation

To prepare our thematic map visualising the distribution of Singapore residents of various attributes, we need to filter the population dataset to only include values from the year 2020.

The `%>%` operator is used extensively here, so it might be worth summarising

```{r}
population_2020 <- population_data %>%
  filter(Time == 2020) %>%

```