---
title: "HX6: Spatial Autocorrelation"
subtitle: "Hands-On Exercise 6"
author: "Kendrick Teo"
date: "2024-09-21"
date-modified: "last-modified"

execute: 
  eval: true
  echo: true
  freeze: true
---

## HX6.1 The analytical question

In spatial policy, a development objective of local government and planners is to ensure **equal distribution of development**. The appropriate spatial statistical methods can be applied to discover if development is evenly distributed, and if not, to find out if and where there are signs of *spatial clustering*.

Today, we return to [Hunan](https://en.wikipedia.org/wiki/Hunan) province in the People's Republic of China to examine the spatial distribution of **GDP per capita**. Just like the previous exercise, we will import the Hunan province administrative boundaries, and 2012 development indicators.

### HX6.2 Importing packages and data

### HX6.2.1 Importing packages

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse)
```

### HX6.2.2 Importing geospatial data

```{r}
hunan <- st_read(dsn="data/geospatial", layer="Hunan")
```

### HX6.2.3 Importing aspatial data

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

### HX6.2.4 Relational join

Here, we will **left join** the attribute fields of the Hunan dataframe with the geographical boundaries of Hunan province.

```{r}
hunan <- left_join(hunan, hunan2012) %>% select(1:4, 7, 15)
```

## HX6.3 Basic visualisation

Before performing any analysis, it is a good practice to create some basic visualisations.

```{r}
equal <- tm_shape(hunan) +
  tm_fill("GDPPC",
          n = 5,
          style = "equal") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Equal interval classification")

quantile <- tm_shape(hunan) +
  tm_fill("GDPPC",
          n = 5,
          style = "quantile") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Equal quantile classification")

tmap_arrange(equal, 
             quantile, 
             asp=1, 
             ncol=2)
```

## HX6.4 Global Measures of Spatial Autocorrelation

Before computing any global spatial autocorrelation statistics, we first need to construct the spatial weights of the study area. These are used to define the neighbourhood relationships between the geographical units in the study area.

### HX6.4.1 Contiguity Spatial Weights

In the code chunk below, `poly2nb()` from the `spdep` pckage is used to compute contiguity weight matrices for the study area. Today, we will create a QUEEN contiguity weight matrix.

```{r}
wm_q <- poly2nb(hunan, 
                queen=TRUE)
summary(wm_q)
```

### HX6.4.2 Creating Row-standardised Weights Matrix

Following this, we need to assign weights to each neighbouring polygon. In our case, we will assign each neighbouring polygon an equal weight, by taking the inverse of the number of neighbours, using `style="W"` for simplicity.

> More options, such as `style="B"`, are available.

```{r}
rswm_q <- nb2listw(wm_q, style="W", zero.policy = TRUE)
rswm_q
```

## HX6.5 Moran's I

**Moran's I** describes how features differ from the values in the study area as a whole. Positive (`I > 0`) values indicate clustering, with observations tending to be similar, while negative (`I` \< 0) values indicate dispersion, where observations tend to be dissimilar. When `I` is approximately 0, it indicates observations are randomly arranged in space. Moran's I approaches `1` when similar values are clustered, and approaches `-1` when dissimilar values tend to cluster.

Moran's I statistics testing can be conducted using `moran.test()` from the `spdep` package.

```{r}
moran.test(hunan$GDPPC, 
           listw=rswm_q, 
           zero.policy = TRUE, 
           na.action=na.omit)
```

The Moran's I statistic is approximately `0.3007`, which indicates a tendency for similar values to be clustered together in space. This means regions with high GDP per capita are likely to be near other similar regions, and vice-versa.

Furthermore, the standard deviate is `4.74` and the p-value is `1.095e-06`. We can reject the null hypothesis that there is no spatial autocorrelation, and since there is significant evidence of positive spatial autocorrelation, we accept this alternative hypothesis.

### HX6.5.1 Monte Carlo on Moran's I

We can perform a Monte Carlo permutation test for the Moran's I statistic using `moran.mc()` and visualise the results.

```{r}
set.seed(1234)
bperm= moran.mc(hunan$GDPPC, 
                listw=rswm_q, 
                nsim=999, 
                zero.policy = TRUE, 
                na.action=na.omit)
bperm

summary(bperm$res[1:999])
hist(bperm$res, 
     freq=TRUE, 
     breaks=20, 
     xlab="Simulated Moran's I")
abline(v=0, 
       col="red") 
```

The Moran's I statistic is `0.30075`. Since the p-value is `0.001`, this shows that the result is statistically significant. Since we have a slightly left-skewed distribution of simulated values, this high level of clustering is unlikely to occur by chance.

## HX6.6 Geary's C

**Geary's C** describes how features differ from their immediate neighbours. When `C > 1`, the points are dispersed, and while `C < 1`, the points are clustered. A `C` value of `1` indicates observations are randomly arranged in space. **Geary's C** approaches `0` when similar values are clustered, and approaches `3` when dissimilar values are clustered. Geary's C is inversely proportional to Moran's I, so higher `C`s correspond to lower `I`s.

Geary's C testing can be performed by calling `geary.test()`, also from `spdedp`.

```{r}
geary.test(hunan$GDPPC, listw=rswm_q)
```

The C statistic is `0.6907`, which suggests positive spatial autocorrelation. With a very low p-value of `0.0001`, we can reject the null hypothesis that there is no spatial autocorrelation.

### HX6.6.1 Monte Carlo on Geary's C

Similar to Moran's I, we can perform a Monte Carlo permutation test on Geary's C and visualise the results.

```{r}
set.seed(1234)
bperm=geary.mc(hunan$GDPPC, 
               listw=rswm_q, 
               nsim=999)
bperm
summary(bperm$res[1:999])
hist(bperm$res, freq=TRUE, breaks=20, xlab="Simulated Geary c")
abline(v=1, col="red") 
```

The distribution of the simulated values indicates a high tendancy towards C-values of 1 or greater, indicating a tendency towards no or negative spatial autocorrelation under random correlations.

## HX6.7 Spatial Correlograms

Spatial correlograms help us examine patterns of spatial autocorrelation. We can create correlograms for both Moran's I and Geary's C.

### HX6.7.1 Moran's I

```{r}
MI_corr <- sp.correlogram(wm_q, 
                          hunan$GDPPC, 
                          order=6, 
                          method="I", 
                          style="W")
plot(MI_corr)
print(MI_corr)
```

### HX6.7.2 Geary's C

```{r}
GC_corr <- sp.correlogram(wm_q, 
                          hunan$GDPPC, 
                          order=6, 
                          method="C", 
                          style="W")
plot(GC_corr)
print(GC_corr)
```

## HX6.8 Local Measures of Spatial Autocorrelation

**Local measures of spatial autocorrelation (LMSA)** focuses on the relationships between each observation and its surroundings, unlike global measures of spatial autocorrelation where a single summary of these relationships is investigated across the map. In this sense, LMSA does not generate summary statistics, but rather, scores that allow us to learn more about the saptial structure in our data.

#### HX6.8.0.1 Local Indicators of Spatial Association

**Local indicators of spatial association (LISA)** are statistics that evaluate the existence of clusters and/or outliers in the spatial arrangement of a given variable. If we are studying GDP per capita, for instance, local clusters in GDP per capita mean that there are areas that have higher or lower rates than is to be expected by chance alone; that is, the values occurring are above or below those of a random distribution in space.

### HX6.8.1 Computing Local Moran's I

The Moran's I statistic can be calculated locally, giving us `li` values given a set of `zi` values and a `listw` object providing neighbour weighting information for the polygon associated with the `zi` values.

We can compute and list the values of the derived Moran's I matrix using the code chunk below.

```{r}
fips <- order(hunan$County)
localMI <- localmoran(hunan$GDPPC, rswm_q)
head(localMI)

printCoefmat(data.frame(
  localMI[fips,], 
  row.names=hunan$County[fips]),
  check.names=FALSE)
```

### HX6.8.1 Mapping Local Moran's I

Once we have our matrix, we are able to plot these values on a `tmap`. Before mapping the Moran's I matrix, it is wise to append the dataframe onto the Hunan polygon featureset.

```{r}
hunan.localMI <- cbind(hunan,localMI) %>%
  rename(Pr.Ii = Pr.z....E.Ii..)
tm_shape(hunan.localMI) +
  tm_fill(col = "Ii", 
          style = "pretty",
          palette = "RdBu",
          title = "local moran statistics") +
  tm_borders(alpha = 0.5)
```

#### HX6.8.1.1 Mapping p-values

The choropleth map shows evidence of both positive and negative `Ii` values. However, it is useful to also consider the p-values for each of these values.

```{r}
tm_shape(hunan.localMI) +
  tm_fill(col = "Pr.Ii", 
          breaks=c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
          palette="-Blues", 
          title = "local Moran's I p-values") +
  tm_borders(alpha = 0.5)
```

#### HX6.8.1.2 Plotting them together

We can also plot both maps side by side with the following code chunk.

```{r}
localMI.map <- tm_shape(hunan.localMI) +
  tm_fill(col = "Ii", 
          style = "pretty", 
          title = "local moran statistics") +
  tm_borders(alpha = 0.5)

pvalue.map <- tm_shape(hunan.localMI) +
  tm_fill(col = "Pr.Ii", 
          breaks=c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
          palette="-Blues", 
          title = "local Moran's I p-values") +
  tm_borders(alpha = 0.5)

tmap_arrange(localMI.map, pvalue.map, asp=1, ncol=2)
```

## HX6.9 Creating a LISA cluster map

LISA cluster maps show significant locations colour-coded by type of spatial autocorrelation.

### HX6.9.1 Plotting Moran scatterplot

Before creating a LISA cluster map, we should first create a Moran scatterplot. This scatterplot illustrates the relationship between the values of our chosen attribute at each location and the average value of the same attribute at neighbouring locations. We can plot a Moran scatterplot for **GDP per capita in 2012** using the `moran.plot()` from `spdep`.

```{r}
nci <- moran.plot(hunan$GDPPC, rswm_q,
                  labels=as.character(hunan$County), 
                  xlab="GDPPC 2012", 
                  ylab="Spatially Lagged GDPPC 2012")
```

Notice that the plot is split in 4 quadrants. The top right corner corresponds to areas that have high GDP per capita and are surrounded by other areas that have the average level of GDP per capita (i.e. positive autocorrelation - clustering).

### HX6.9.2 Plotting Moran scatterplot with standardised variable

We can use `scale()` to center and scale our variable. Here centering is done by subtracting the mean (omitting NAs) the corresponding columns, and scaling is done by dividing the (centered) variable by their standard deviations.

```{r}
hunan$Z.GDPPC <- scale(hunan$GDPPC) %>% 
  as.vector 
# Ensures we get a vector that maps neatly into our dataframe

nci2 <- moran.plot(hunan$Z.GDPPC, rswm_q,
                   labels=as.character(hunan$County),
                   xlab="z-GDPPC 2012", 
                   ylab="Spatially Lag z-GDPPC 2012")
```

### HX6.9.3 Preparing LISA map classes

The code chunks below showcase the steps to prepare a LISA cluster map.

```{r}         
quadrant <- vector(mode="numeric",length=nrow(localMI))
```

Next, we derive the spatially lagged variable of interest (i.e. `GDPPC`) and centers the spatially lagged variable around its mean.

```{r}      
hunan$lag_GDPPC <- lag.listw(rswm_q, hunan$GDPPC)
DV <- hunan$lag_GDPPC - mean(hunan$lag_GDPPC)     
```

This is followed by centering the local Moran’s around the mean.

```{r}      
LM_I <- localMI[,1] - mean(localMI[,1])    
```

Next, we set a statistical significance level for the local Moran.

```{r}      
signif <- 0.05       
```

These four lines define the low-low (`1`), low-high (`2`), high-low (`3`) and high-high (`4`) categories.

```{r}      
quadrant[DV <0 & LM_I>0] <- 1
quadrant[DV >0 & LM_I<0] <- 2
quadrant[DV <0 & LM_I<0] <- 3  
quadrant[DV >0 & LM_I>0] <- 4      
```

Lastly, we place the non-significant Moran in the category 0.

```{r}      
quadrant[localMI[,5]>signif] <- 0
```

### HX6.9.4 Plotting LISA map

Finally, we can build our LISA map as below.

```{r}
hunan.localMI$quadrant <- quadrant
colors <- c("#ffffff", "#2c7bb6", "#abd9e9", "#fdae61", "#d7191c")
clusters <- c("insignificant", "low-low", "low-high", "high-low", "high-high")

tm_shape(hunan.localMI) +
  tm_fill(col = "quadrant", 
          style = "cat", 
          palette = colors[c(sort(unique(quadrant)))+1], 
          labels = clusters[c(sort(unique(quadrant)))+1],
          popup.vars = c("")) +
  tm_view(set.zoom.limits = c(11,17)) +
  tm_borders(alpha=0.5)
```

For a more effective interpretation, it is good practice to plot our basemap, local Moran's I values map and p-value map side by side.

```{r}
gdppc <- qtm(hunan, "GDPPC")

hunan.localMI$quadrant <- quadrant
colors <- c("#ffffff", "#2c7bb6", "#abd9e9", "#fdae61", "#d7191c")
clusters <- c("insignificant", "low-low", "low-high", "high-low", "high-high")

LISAmap <- tm_shape(hunan.localMI) +
  tm_fill(col = "quadrant", 
          style = "cat", 
          palette = colors[c(sort(unique(quadrant)))+1], 
          labels = clusters[c(sort(unique(quadrant)))+1],
          popup.vars = c("")) +
  tm_view(set.zoom.limits = c(11,17)) +
  tm_borders(alpha=0.5)

tmap_arrange(gdppc, LISAmap, 
             asp=1, ncol=2)
```

## HX6.10 Hot Spot and Cold Spot Areas Analysis

Localised spatial statistics can also be used to detect **hot spot** and/or **cold spot** areas. One way to do this is by creating Getis and Ord's G-Statistics. It looks at neighbours within a defined proximity to identify whether either high or low values cluster spatially. The analysis consists of three steps:

1. Deriving spatial weight matrix
2. Computing Gi statistics
3. Mapping Gi statistics

### HX6.10.1 Deriving distance-based weight matrix

First, we need to define a new set of neighbours based on distance. There are two types of distance-based proximity matrices: fixed-distance, and adaptive-distance.

#### HX6.10.1.1 Deriving the centroid

We will need points to associate with each polygon before we can create our connectivity graph. It will be a little more complicated than just running `st_centroid()` on the sf object; we need the coordinates in a separate data frame for this to work. To do this we will use a mapping function. The mapping function applies a given function to each element of a vector and returns a vector of the same length. Our input vector will be the geometry column of us.bound. Our function will be `st_centroid()`.

To get our longitude values we map the st_centroid() function over the geometry column of us.bound and access the longitude value through double bracket notation [[]] and 1. This allows us to get only the longitude, which is the first value in each centroid.

```{r}
longitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[1]])
```

We do the same for latitude with one key difference. We access the second value per each centroid with `[[2]]`.

```{r}
latitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[2]])
```

Now that we have latitude and longitude, we can use `cbind` to put longitude and latitude into the same object.

```{r}
coords <- cbind(longitude, latitude)
```

#### HX6.10.1.2 Determining the cutoff distance

We now need to determine the upper limit for our distance band.

```{r}
k1 <- knn2nb(knearneigh(coords))
k1dists <- unlist(nbdists(k1, coords, longlat = TRUE))
summary(k1dists)
```

The summary report shows that the largest first nearest neighbour distance is 61.79 km, so using this as the upper threshold gives certainty that all units will have at least one neighbour.

#### HX6.10.1.3 Computing fixed-distance weight matrix

We can now compute our distance weight matrix with `dnearneigh()`, and spatial weights object with `nb2listw()`.

```{r}
wm_d62 <- dnearneigh(coords, 0, 62, longlat = TRUE)
wm_d62
wm62_lw <- nb2listw(wm_d62, style = 'B')
summary(wm62_lw)
```

### HX6.10.2 Computing adaptive distance weight matrix

A fixed distance weight matrix can cause more densely settled areas (usually urban areas) to have more neighbours, and less densly settled areas (usually rural areas) to have fewer neighbours. Having many neighbours smoothes the neighbour relationship across more neighbours.

It is possible to control the numbers of neighbours directly using k-nearest neighbours, either accepting asymmetric neighbours or imposing symmetry.

```{r}
knn <- knn2nb(knearneigh(coords, k=8))
knn
knn_lw <- nb2listw(knn, style = 'B')
summary(knn_lw)
```

### HX6.10.3 Computing Gi statistics

Gi-statistics show us where features with either high (i.e. hot spots) or low (i.e. cold spots) cluster spatially. **Hot spot areas** are areas that are significant and positive if location i is associated with relatively high values of the surrounding locations, while **cold spot areas** are areas that are significant and negative if location i is associated with relatively low values in surrounding locations.

#### HX6.10.3.1 Gi statistics using fixed distance

```{r}
fips <- order(hunan$County)
gi.fixed <- localG(hunan$GDPPC, wm62_lw)
gi.fixed
```

The Gi statistics are represented as a Z-score. Higher Z-scores represent a greater intensity of clustering, while the direction (positive or negative) indicates high or low clusters.

Now, we will join the Gi values to the `hunan` dataframe.

```{r}
hunan.gi <- cbind(hunan, as.matrix(gi.fixed)) %>%
  rename(gstat_fixed = as.matrix.gi.fixed.)
```

#### HX6.10.3.2 Mapping Gi values with fixed distance weights

```{r}
gdppc <- qtm(hunan, "GDPPC")

Gimap <-tm_shape(hunan.gi) +
  tm_fill(col = "gstat_fixed", 
          style = "pretty",
          palette="-RdBu",
          title = "local Gi") +
  tm_borders(alpha = 0.5)

tmap_arrange(gdppc, Gimap, asp=1, ncol=2)
```

#### HX6.10.3.3 Gi statistics with adaptive distance

```{r}
fips <- order(hunan$County)
gi.adaptive <- localG(hunan$GDPPC, knn_lw)
hunan.gi <- cbind(hunan, as.matrix(gi.adaptive)) %>%
  rename(gstat_adaptive = as.matrix.gi.adaptive.)
```

#### HX6.10.3.4 Mapping Gi values with adaptive distance weights

```{r}
gdppc<- qtm(hunan, "GDPPC")

Gimap <- tm_shape(hunan.gi) + 
  tm_fill(col = "gstat_adaptive", 
          style = "pretty", 
          palette="-RdBu", 
          title = "local Gi") + 
  tm_borders(alpha = 0.5)

tmap_arrange(gdppc, 
             Gimap, 
             asp=1, 
             ncol=2)
```

The hot spot areas are the high GDP per capta areas clustered towards the northeast of the study region, while the cold spot areas are clustered towards the south and southwest.