---
title: "CE10: Geographically Weighted Regression"
subtitle: "In-class Exercise 10"
author: "Kendrick Teo"
date: "2024-10-28"
date-modified: "last-modified"

execute: 
  eval: true
  echo: true
  freeze: true
---

## CE10.1 Getting Started

```{r}
pacman::p_load(olsrr, ggstatsplot, sf, tmap, tidyverse, gtsummary, performance, see, sfdep)

mpsz <- st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_WEB_PL")
condo_resale <- read_csv("data/aspatial/Condo_resale_2015.csv")
condo_resale_sf <- st_as_sf(condo_resale, coords = c("LONGITUDE", "LATITUDE"), crs=4326) %>% st_transform(crs=3414)
```

## CE10.2 Correlation Analysis with `ggstatplot`

```{r}
ggcorrmat(condo_resale[, 5:23])
```

## CE10.3 Multiple Linear Regression method

```{r}
condo_mlr <- lm(formula = SELLING_PRICE ~ AREA_SQM + 
                  AGE   + PROX_CBD + PROX_CHILDCARE + 
                  PROX_ELDERLYCARE + PROX_URA_GROWTH_AREA + 
                  PROX_HAWKER_MARKET    + PROX_KINDERGARTEN + 
                  PROX_MRT  + PROX_PARK + PROX_PRIMARY_SCH + 
                  PROX_TOP_PRIMARY_SCH + PROX_SHOPPING_MALL + 
                  PROX_SUPERMARKET + PROX_BUS_STOP + 
                  NO_Of_UNITS + FAMILY_FRIENDLY + 
                  FREEHOLD + LEASEHOLD_99YR, 
                data=condo_resale_sf)
summary(condo_mlr)
```

## CE10.4 Tidy linear regression report with OLS

```{r}
ols_regress(condo_mlr)
```

## CE10.5 Stewpise regression

Stepwise regression involves constructing a regression model by selecting independent variables step by step. It involves adding or removing potential explanatory variables in succession and testing for statistical significance after each iteration. Three methods are available: `ols_step_forward_p`, `ols_step_backward_p`, and `ols_step_both_p`.

### CE10.5.1 Forward stepwise

```{r}
condo_fw_mlr <- ols_step_forward_p(
  condo_mlr,
  p_val = 0.05,
  details = FALSE)
condo_fw_mlr
plot(condo_fw_mlr)
```

### CE10.5.2 Backward stepwise

```{r}
condo_bw_mlr <- ols_step_backward_p(
  condo_mlr,
  p_val = 0.05,
  details = FALSE)
condo_fw_mlr
plot(condo_fw_mlr)
```

### CE10.5.3 Stepwise

```{r}
condo_sb_mlr <- ols_step_both_p(
  condo_mlr,
  p_val = 0.05,
  details = FALSE)
condo_sb_mlr
```

## CE10.6 Comparing performance

```{r}
metric <- compare_performance(condo_mlr, 
                    condo_fw_mlr$model,
                    condo_bw_mlr$model,
                    condo_sb_mlr$model)

metric$Name <- gsub(".*\\\\([a-zA-Z0-9_]+)\\\\, \\\\model\\\\.*", "\\1", metric$Name)
```

## CE10.7 Visualising model parameters

```{r}
ggcoefstats(condo_sb_mlr$model,
            sort = "ascending")
```

## CE10.8 Checking for multilinearity

```{r}
check_collinearity(condo_sb_mlr$model)

plot(check_collinearity(condo_sb_mlr$model)) +
  theme(axis.text.x = element_text(
    angle = 45, hjust = 1))
```

## CE10.9 Linearity assumption test

```{r}
out <- plot(check_model(condo_sb_mlr$model, 
                        panel = FALSE))
out[[2]]
```

## CE10.10 Normality assumption test

```{r}
plot(check_normality(condo_sb_mlr$model))
```

## CE10.11 Checking of outliers

```{r}
outliers <- check_outliers(condo_sb_mlr$model,
                           method = "cook")
outliers
```

## CE10.12 Visualising spatial non-stationary

```{r}
mlr_output <- as.data.frame(condo_fw_mlr$model$residuals) %>%
  rename(`FW_MLR_RES` = `condo_fw_mlr$model$residuals`)

condo_resale_sf <- cbind(condo_resale_sf, 
                        mlr_output$FW_MLR_RES) %>%
  rename(`MLR_RES` = `mlr_output.FW_MLR_RES`)

tmap_mode("view")
tm_shape(mpsz)+
  tmap_options(check.and.fix = TRUE) +
  tm_polygons(alpha = 0.4) +
tm_shape(condo_resale_sf) +  
  tm_dots(col = "MLR_RES",
          alpha = 0.6,
          style="quantile") 
tmap_mode("plot")
```

## CE10.13 Spatial stationary test (Moran's I)

```{r}
condo_resale_sf <- condo_resale_sf %>%
  mutate(nb = st_knn(geometry, k=6,
                     longlat = FALSE),
         wt = st_weights(nb,
                         style = "W"),
         .before = 1)

global_moran_perm(condo_resale_sf$MLR_RES, 
                  condo_resale_sf$nb, 
                  condo_resale_sf$wt, 
                  alternative = "two.sided", 
                  nsim = 99)
```
